My first Notebook 

```{r}
plot(cars)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

Load libary ggplot 2
```{r}
library(ggplot2)
```
Reads the gapMinderDataFiveYear data set into an object
```{r}
data.pull <-read.csv("gapMinderDataFiveYear.csv")
```
Create a summary of the data
```{r}
summary(data.pull)
```
plot plot of population vs life expectancy
```{r}
plot(data.pull[,"pop"],data.pull[,"lifeExp"], main = "Pop vs. LifeExp",xlab="Pop", ylab="LifeExp")
```
Creates a plot of the distribution of population using the geom_histogram function
from the ggplot2 package
```{r}
ggplot(data.pull,aes(x=data.pull[,"pop"])) +
  geom_histogram() +
  labs(title="Histogram for Pop", x="Pop", y="Count") +
  xlim(c(6.001e+04,1.319e+09)) +
  ylim(c(0,260))
```



