---
title: 'Module 1.3 Predictive Analytics Tools: Introduction to R'
output:
  html_notebook: default
---

Run CHUNK 1 to see some examples of R syntax.

```{r}
#CHUNK 1
x <- 1
y = 5

x + y

a <- "Hello"
b = "World"

paste(a,b)
```

CHUNK 2 illustrates the difference between <- and =.

```{r}
#CHUNK 2
# Use "<-" to place a value into a variable, use "=" when naming arguments to a function
x <- runif(10, min = -10, max = 10)
x
```

CHUNK 3 illustrates the use of comments.

```{r}
#CHUNK 3
# This is a script to illustrate comments

###
# The following code adds x and y
# and prints whether the result is
# greater than 5
###

x <- runif(1,min=0,max=5) # x is uniform distributed between 0 and 5
y <- runif(1,min=0,max=7) # y is uniform distributed between 0 and 7

if (x + y > 5) {
	# You can indent comments to make the code neater
	print("result is greater than 5")
} else {
  print ("result is less than 5")
}
```

Run CHUNK 4 to see how a function is written. 

```{r}
#CHUNK 4
# The function name must be preceeded with "function"
MyFunction <- function(argument1, argument2, argument3 = 0) { # The arguments (and default values if any) 
                                                              # are supplied within parentheses and 
                                                              # separated by a comma
  
	# The body of the function is inside a pair of brackets: {}
	string <- paste("Argument1 is",argument1,". Argument2 is",argument2, ". Argument3 is",argument3,".")
	print("MyFunction did something.")
	
	# To explicitly return a value from the function use "return"
	return(string)
}

# You can use your function in exactly the same way as all other R functions, just specify its name and 
# the arguments (make sure you run the code above to define your function first or the program won't know it exists).
MyFunction("hello","blue","three")

# If the function has a return value you can place it into a variable
result <- MyFunction(1,2)
result
```

Fill in CHUNK 5 to write your own function to complete Exercise 1.3.5.

```{r}
#CHUNK 5
# This function takes two arrays, a1 and a2, as arguments and returns the absolute difference between: 
#   - The sum of squares of the elements of a1 (e.g. 1^2 + 3^2 + 7^2); and
#   - The square of the sum of the elements of a2 (e.g. (2 + 5)^2)

##
# Write your function definition here
##

# Test your function with the following code: all statements should return TRUE
# Note you can adjust the name of the function below to match what you decide to call your function
diffSquares(c(1, 2, 3), c(2, 3, 4)) == 67
diffSquares(c(3, 4), c(0.5, -1)) == 24.75
diffSquares(a2 = c(2:10), a1 = c(-5:5)) == 2806
diffSquares(c(1:100),c(1:100))

``` 

CHUNK 6 contains a sample solution.

```{r}
#CHUNK 6
diffSquares <- function(a1 = c(), a2 = c()) {
  if (length(a1) == 0 || length(a2) == 0) {
    stop("Both arguments must have a length of at least one.")
  }
  sumSquares <- sum(a1^2)
  squareSum <- (sum(a2))^2
  return(abs(sumSquares - squareSum))
}

diffSquares(c(1, 2, 3), c(2, 3, 4)) == 67
diffSquares(c(3, 4), c(0.5, -1)) == 24.75
diffSquares(a2 = c(2:10), a1 = c(-5:5)) == 2806
diffSquares(c(1:100),c(1:100))
```

Run CHUNK 7 to create several data frames.

```{r}
#CHUNK 7
# Create a data frame from nothing
df.empty <- data.frame(column1 = numeric(), column2 = character(), column3 = factor())
df.empty$column3<-factor(df.empty$column3,levels=c("A","B","C"))
df.manual <- data.frame(x = c(1,2,3), y = c("Amy", "John","Cameron"), z = c("A","B","C"))

# Create a data frame from a matrix
m <- matrix(data = c(1,2,3,4,5,6), nrow = 2, ncol = 3)
df.matrix <- as.data.frame(m)

# Create a data frame from data
df.data <- read.csv("BreastCancerWisconsinDataSet.csv")

df.empty
df.manual
df.matrix
df.data
```

CHUNK 8 accesses data from one of the previously created data frames.

```{r}
#CHUNK 8
# Direct indexing (note our data frame has two dimensions so we index both dimensions using a comma to separate)
df.manual[,3]
df.manual[1,3]
df.manual[1,]
df.manual[1:3,c(1,3)]

# Using column names (or row names if you have specified them) -- This is preferred to direct indexing where possible
df.manual[,'x']
df.manual[2,c('x','z')]
df.manual[which(df.manual[,'y']=="John"),c('x','y','z')]

# Columns are considered "attributes" of R data frames, so we can use the "attribute" accessor "$"
df.manual$x
df.manual$z[2] 
df.manual$y[which.max(df.manual$x)]

# Note by accessing the column first ($z and $y) we have reduced the object to one 
# dimension (just the z or y column) so no comma required when indexing
```

CHUNK 9 provides several examples of manipulating data frames. Some of the code has been designed to generate error messages. You will need to eliminate them to run subsequent lines.

```{r}
#CHUNK 9
# Adding rows (note data types must be consistent)
newRow <- c(-2,"MyString","B")
incorrectRow <- c(-2,"MyString","D")
df.empty <- rbind(df.empty,newRow)
df.empty
df.empty <- rbind(df.empty,incorrectRow)
df.empty

# Adding columns (note must have the same number of rows)
newColumn <- runif(3)
incorrectColumn <-  runif(5)
df.manual <- cbind(df.manual, newColumn)
df.manual
df.manual <- cbind(df.manual, incorrectColumn) # This will fail
 
df.manual$d # creates a new NULL column
df.manual$d <- newColumn

newColumnName <- "anotherColumn"
df.manual[,newColumnName] <- newColumn
df.manual

# Modifying data (just access the elements and replace with something the same size and type)
df.manual$d[1:2] <- runif(2)
df.manual
is.character(df.manual) # You can also access elements using TRUE/FALSE - this is useful when checking for NA values using the function is.na()
df.manual[is.character(df.manual)] <- "Z"
df.manual

# Replacing values using conditions
df.manual[, "anotherColumn"] <- ifelse(df.manual$y == "John", 1, 0)
df.manual

# Changing the column names
colnames(df.manual)
colnames(df.manual) <- c("Ant","Bee","Cat","Dog","Elephant","Giraffe")
df.manual
```

CHUNK 10 provides space for completing Exercise 1.3.6.

```{r}
#CHUNK 10
# Write your code here


```

CHUNK 11 provides a sample solution.

```{r}
#CHUNK 11
# 1)
data.all <- read.csv("gapMinderDataFiveYear.csv",stringsAsFactors = TRUE)

# 2)
data.bin <- data.all # create the new data frame

# Iterate through each level in the factor "continent"
for (level in levels(data.bin$continent)) {
  # Create the name of the new variable (e.g. continent_Europe)
  varName <- paste("continent_",level,sep = "") 
  
  # Create the new column, if the value in "continent" is equal to the current level, then 1, otherwise 0.
  data.bin[, varName] <- ifelse(data.bin$continent == level, 
                                1,
                                0)
}

# 3)
Binarize <- function(dataframe, variableList) {
  # For each factor variable in variableList, creates binary flag columns in df for each level in those variables
  
  for (variable in variableList) {
    var.column <- dataframe[,variable]
    
      if (!(is.factor(var.column))) {
        warning(paste(variable, "variable is not of type factor, not binarized\n"))
        }
    
    
    for (level in levels(var.column)) {
      varName <- paste(variable, "_", level, sep = "") 
      
      dataframe[, varName] <- ifelse(var.column == level, 
                                1,
                                0)
    }
  }
  return(dataframe)
}
```

CHUNK 12 provides space for your solution to Exercise 1.3.7.

```{r}
#CHUNK 12
# Write your code here



```

CHUNK 13 provides a sample solution.

```{r}
#CHUNK 13
# 1)

AppendResults <- function(resultsTable = NULL, newModelName, predicted, actual) {
  
  ## If we don't already have a results table then create an empty one
  if (is.null(resultsTable)) {
    resultsTable <- data.frame(modelName = character(0), errorRate = numeric(0))
  }
  
  ## Calculate the result:
  error = mean(abs(predicted - actual))
  
  ## Append results to table
  newrow <- data.frame(modelName = newModelName, errorRate = error)
  resultsTable <- rbind(resultsTable, newrow)
  
  return(resultsTable)
}

# 2)
## Assuming we have the results still in memory (re-run Section 1.1 to get them)
results <- AppendResults(newModelName = "Logistic_Regression", predicted = Logistic.m1.pred.test, actual = data.testing$target)
results <- AppendResults(resultsTable = results, newModelName = "Decision_Tree", predicted = Tree.m1.pred.test, actual = data.testing$target)

# 3)
randomModel.pred <- rbinom(n = nrow(data.testing), size = 1, prob = (sum(data.training$target))/nrow(data.training))
results <- AppendResults(resultsTable = results, newModelName = "Random", predicted = randomModel.pred, actual = data.testing$target)

results
```

For Part 1 of Exercise 1.3.9 examine CHUNK 14 and determine what it does. Then modify the Plot.Dis function as instructed in Part 2.

```{r}
#CHUNK 14
# Load required packages
required.packages = c("ggplot2", "readr")
invisible(lapply (required.packages, require, character.only=TRUE))

# Load data 
data.all <- read.csv("BreastCancerWisconsinDataSet.csv")

# Basic cleaning
 # set the target: we are trying to predict whether a diagnosis is malignant (M) or benign (B).
data.all$target [data.all$diagnosis == "M"] = 1
data.all$target [data.all$diagnosis == "B"] = 0

 # all variables available for training
vars <- names(data.all)[c(-1, -2, -33)]
data.all <- data.all[c(-1, -2, -33)]

# Function for creating a certain plot given some data as input
Plot.Dis <- function (data) {
  ggplot(data, aes(x = data$target)) + 
  geom_bar(aes(fill = data$target, color = data$target)) + 
  ggtitle(paste("Distribution of target for", 
                deparse(substitute(data)), sep = " ")) + 
  theme(legend.position = "none")
}

Plot.Dis(data.all)
```

CHUNK 15 provides a sample solution.

```{r}
#CHUNK 15
# 2.a)
Plot.Dis <- function (data) {
  ggplot(data, aes(x = radius_mean)) + 
  geom_bar(aes(fill = radius_mean, color = radius_mean)) + 
  ggtitle(paste("Distribution of radius_mean for", 
                deparse(substitute(data)), sep = " ")) + 
  theme(legend.position = "none")
}

Plot.Dis(data.all)

# 2.b)
Plot.Dis <- function (data, var) {
  ggplot(data, aes(x = data[, var])) + 
  geom_bar(aes(fill = data[, var], color = data[, var])) + 
  ggtitle(paste("Distribution of var for", 
                deparse(substitute(data)), sep = " ")) + 
  theme(legend.position = "none")
}

Plot.Dis(data.all, "radius_mean")
```

Modify CHUNK 16 per the instructions in Exercise 1.3.10.

```{r}
#CHUNK 16
numSeq<-c(0:9)
n <- 1e6
GetNthLexPermutation <- function(sequence,n) {len <- length(sequence)  
  if (n > prod(1:len)|n<1) {stop("The number required is beyond the number of possible permutations.")}
  else {nth <- numeric(len)  
    for (i in 1:(len - 1)) {
      if (n==0) {nth[i]<-sequence[length(sequence)]}
      else {tmp<-prod(1:(len-i)) 
        div<-floor((n-1)/tmp)
        nth[i]<-sequence[div+1]}
      n <- n%%tmp
      sequence <-sequence[sequence!=nth[i]]}
    nth[len] <- sequence
    return(nth)}} 
cat("The result is:",GetNthLexPermutation(numSeq,n),"\n")
```

The solution to Exercise 1.3.10 is in CHUNK 17.

```{r}
#CHUNK 17
numSeq <- c(0:9)
n <- 1e6

# Gets the nth lexicographical permutation of the given sequence
GetNthLexPermutation <- function(sequence, n) {
  
  len <- length(sequence)  # calculate the length of the sequence
  
  # Check if n is within the total number of permutations
  if (n > prod(1:len) | n < 1) {
    stop("The number required is outside the number of possible permutations.")
  } else {
    nth <- numeric(len)  # Create a vector called nth of length len to store the nth permutation
    
    # Go through each place in the vector nth and work out what character belongs there
    for (i in 1:(len - 1)) {
      # We change n each time, when it is 0 we want the last character in what is left of the sequence
      if (n == 0) { 
        nth[i] <- sequence[length(sequence)]
      } else {
        # The next character is given by the (n-1)/(total number of permutations) position in the remaining sequence
        tmp <- prod(1:(len - i)) 
        div <- floor((n - 1) / tmp)
        nth[i] <- sequence[div + 1]
      }
      
      n <- n %% tmp  # Change n to be n mod tmp (we have found the first few characters that start
                     # the nth permutation, so we are looking for the remaining characters which is the (n mod tmp)th
                     # permutation of the remaining characters)
      sequence <- sequence[sequence != nth[i]]  # Remove the character we just used from the sequence 
    }
    
    nth[len] <- sequence  # There is only one character left in the sequence which is the last one
    return (nth)  # Return the sequence
  }
}

# Print out the results
cat("The result is:", GetNthLexPermutation(numSeq, n), "\n")
```

Exercise 1.3.11: Use CHUNK 18 to load the gapminder data into an object.

```{r}
#CHUNK 18
# Read the data into an object called "data.full"

```

CHUNK 19 provides space to complete Exercise 1.3.11.

```{r}
#CHUNK 19
# Write your R code here
```

CHUNK 20 contains a sample solution to Exercise 1.3.11.

```{r}
#CHUNK 20
# Sample solution

# Read the data into an object called data.full:
data.full <- read.csv("gapminderDataFiveYear.csv",stringsAsFactors = TRUE)

# Do some basic exploration
colnames(data.full)
data.full[1:10,]
summary(data.full)
sapply(data.full, class)

```

Use CHUNK 21 to write your code for Exercise 1.3.12.
    
```{r}
#CHUNK 21
# Write your R code here


```

CHUNK 22 has a solution to Exercise 1.3.12.

```{r}
#CHUNK 22
# Necessry functions
Plot.Dis <- function (data) {
  ggplot(data, aes(x = data$lifeExp)) + 
  geom_bar(aes(fill = 0, color = 0),
           stat = "bin") + 
  ggtitle(paste("Distribution of lifeExp for", 
                deparse(substitute(data)), sep = " ")) + 
  theme(legend.position = "none")
}


# Sample solution 1
data.train <- data.full[data.full$year < 2000, ]
data.holdout <- data.full[data.full$year >= 2000, ]

# Sample solution 2
data.full$rand <- runif(nrow(data.full), min = 1, max = 10)
data.train <- data.full[data.full$rand <= 8, ]
data.holdout <- data.full[data.full$rand > 8, ]

# Note these are not the only methods and depending on the problem they might not be the best either.

Plot.Dis(data.full)
Plot.Dis(data.train)
Plot.Dis(data.holdout)
```

CHUNK 23 provides code for saving plots.

```{r eval=FALSE}
#CHUNK 23
png(filename = "PlotDirectoryName/descriptiveNameForPlot.png") # Open the R graphics device
plot(...) # Run your plots
dev.off() # Close the graphics device
```

Use the space in CHUNK 24 to work on Exercise 1.3.13.

```{r}
#CHUNK 24
# Write your R code here


```

CHUNK 25 contains a sample solution to Exercise 1.3.13.

```{r}
#CHUNK 25
# Sample solution
png(filename="plot.png")
plot(data.full$lifeExp~data.full$year, col = data.full$continent)
dev.off()
```

CHUNK 26 provides examples of bad and good code for running a linear model. Don't run this code as the setup is hypothetical, not related to any data set you have loaded.
  
```{r eval = FALSE}
# Bad:

myModel <- glm(Y~A+B+C+D+E+F+G+H+I+J+K+L+M+N+O+P+Q+R+S+T+U+V+W+X+Z,family = gaussian(link = "identity"), control = list(epsilon = 1e-8, maxit = 50, trace = FALSE), data = data.training)

# Good:

# Control parameters
myModel.control1 <- list(epsilon = 1e-8, # Comment about chosen value
                         maxit = 100, 
                         trace = FALSE)
# If we want to test more than one set, define here so easy to see differences
myModel.control2 <- list(epsilon = 1e-8, 
                         maxit = 50, 
                         trace = TRUE)

# GLM Formula
smallformula <- as.formula("Y ~ A + B + C + D + E + F + G")  # Unweildy if their are too many variables

# For a large number of variables we want to build the list of variables we will be modelling 
# as a list first (which we can easily modify if we need to)
model.variables.all <- colnames(data.all)
varsToDrop <- c("Y","C","D","Z")  # specify variables we don't want
model.variables <- model.variables.all[!(model.variables.all %in% varsToDrop)]  # Keep only those variables which 
                                                                                #aren't in the ones we are dropping
bigformula <- as.formula(paste("Y ~",paste(model.variables, collpse = "+"), sep = ""))

# Model 
myModel <- glm(bigformula,
               family = gaussian(link = "identity"),
               control = myModel.control2,
               data = data.training)

```


CHUNK 27 provides space for a solution to Exercise 1.3.14.

```{r}
#CHUNK 27
# Write your code here


```

CHUNK 28 provides a sample solution to Exercise 1.3.14.

```{r}
#CHUNK 28
# Control parameters
glm.control <- list(epsilon = 1e-8,  # Default
                    maxit = 50,  # Default
                    trace = FALSE)  # Default

# GLM formula
model.target <- "lifeExp"
model.variables <- colnames(data.full)[!(colnames(data.full) %in% model.target)]
glm.formula <- as.formula(paste(model.target, " ~ ", paste(model.variables, collapse = "+"), sep = ""))

# GLM model
glm.fit <- glm(glm.formula,
               family = gaussian(link = "identity"),
               control = glm.control,
               data = data.train)


```

CHUNK 29 provides room for your solution to Exercise 1.3.15.

```{r}
#CHUNK 29
# Write your code here


```

CHUNK 30 has a sample solution to Exercise 1.3.15.

```{r}
#CHUNK 30
# Look at coefficiemts
glm.fit$coefficients

# Function for model performance
GetModelPerformance <- function(model, testData, trainData) {
  # Predict both train and test data sets
  model.predict.train <- predict(model, trainData, type = "response")
  model.predict.test <- predict(model, testData, type = "response")
  
  # Get mean error
  error.train <- mean(abs(model.predict.train - trainData[, "lifeExp"]))
  error.test <- mean(abs(model.predict.test - testData[, "lifeExp"]))
  
  # Print results
  print(paste("Training error:", error.train))
  print(paste("Test error:", error.test))
}

GetModelPerformance(glm.fit, data.holdout, data.train)

plot(glm.fit)

```

CHUNK 31 illustrates a command that will save your workspace. 

```{r}
#CHUNK 31
save.image("~/test.RData") # note '~/' refers to the root directory and can be replaced with '' to refer to the current working directory, or  folder path of your choice
```

To reload your workspace later, use the command illustrated in CHUNK 32.

```{r}
#CHUNK 32
load("~/test.RData")
```


CHUNK 33 provides an example of how to save a dataframe for later use. Keep in mind that a path is not needed if everything is stored in your working directory.

```{r}
#CHUNK 33
data.raw <- data.frame(Var1 = c(1,2,3), Var2 = c(4,5,6)) # the raw data

data.transformed <- rbind(data.raw, c(10,11)) # Adds a new row to the data

save(data.transformed, file = "[folder_path]/Module_X_data_transformed.RData") # Note .RData extension 
```

To reload that data in a future R session use code as in CHUNK 34.

```{r}
#CHUNK 34
data.transformed <- load("[folder_path]/Module_X_data_transformed.RData")
```

CHUNK 35 contains code to modify for Exercise 1.3.16.

```{r}
#CHUNK 35
## Load required packages
required.packages = c("ggplot2", "readr", "tidyr", "tree")
lapply (required.packages, require, character.only=TRUE)

## Load data 
data.all <- read.csv("BreastCancerWisconsinDataSet.csv")

## Some simple cleaning and set up
 
# set the target: we are trying to predict whether a diagnosis is malignant (M) or benign (B).
data.all$target [data.all$diagnosis == "M"] = 1
data.all$target [data.all$diagnosis == "B"] = 0

# all variables available for training
vars <- names(data.all)[c(-1, -2, -33)]
data.all <- data.all[c(-1, -2, -33)]

# split data into training vs testing
set.seed(1000)
data.all$rand <- runif(nrow(data.all))
data.training <- data.all[which(data.all$rand <= 0.5), ] 
data.testing <- data.all[which(data.all$rand > 0.5), ]
data.all$rand <- NULL
data.training$rand <- NULL
data.testing$rand <- NULL

##
# Save the data.all, data.training and data.testing objects into appropriately named .RData files
##

## Clean up (dont need data.all anymore)
rm(data.all)

## Fit a simple decision tree model
Tree.m1 <- tree (target ~ radius_mean + 
                           texture_mean +
                           perimeter_mean +
                           area_mean +
                           smoothness_mean	+ 
                           compactness_mean	+
                           concavity_mean	+
                           concave.points_mean +
                           symmetry_mean, 
                 data = data.training)

##
# Save the Tree.m1 model into and appropriately named .RData file
##

## Predict results - re-run this code after you reload the objects you saved. Make sure the results are the same!
Tree.m1.pred.train = predict(Tree.m1, data.training, type = "vector")
Tree.m1.pred.train  = ifelse(Tree.m1.pred.train > 0.5, 1, 0)
error = mean(Tree.m1.pred.train != data.training$target)
print(paste('Tree Training Model Accuracy', 1-error))

Tree.m1.pred.test = predict(Tree.m1, data.testing, type = "vector")
Tree.m1.pred.test = ifelse(Tree.m1.pred.test > 0.5, 1, 0)
error = mean(Tree.m1.pred.test != data.testing$target)
print(paste('Tree Testing Model Accuracy', 1-error))

## Clean up
rm(list = ls()) # This will remove all objects in the environment

```

CHUNK 36 provides code for saving a data file in csv format for use in other programs.

```{r}
#CHUNK 36
write.csv(x = data.transformed,
          file = "[folder_path]/Module_X_data_transformed.csv", # Note .csv extension
          na = "?", # Use to change the value used for NAs (default is "NA")
          row.names = FALSE # Whether you want row names to appear in the first column
)
```

CHUNK 37 provides code for saving a data file in SAS format.

```{r}
#CHUNK 37
library(foreign) # Allows reading and writing of data from/for other programs like SPSS and SAS

write.foreign(dataframe,"[folder_path]/Module_X_data_transformed.sas", package="SAS") # Writes the data as a plain text file and also creates an instruction file that will allow SAS to read it
```

CHUNK 38 illustrates the command for saving your history.

```{r}
#CHUNK 38
savehistory(file = "[folder_path]/Module_X_history.Rhistory")
```

If you wish to load a previously saved history you can use CHUNK 39.

```{r}
#CHUNK 39
loadhistory(file = "[folder_path]/Module_X_history.Rhistory")
```

