---
title: "Module 4.2: Bivariate Data Exploration"
output: html_notebook
---

run CHUNK 1 to be sure ggplot and the SOA mortality data are loaded.

```{r echo = FALSE}
#CHUNK 1
library(ggplot2)
data.mortality <- read.csv("soa_mortality_data.csv")
```

Run CHUNK 2 to create a split box plot.

```{r echo = FALSE}
#CHUNK 2
p1 <- ggplot(data = data.mortality, aes(x = issstate, y = issage)) +
        geom_boxplot(alpha = 0.5) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
p1
```

Run CHUNK 3 to create additional plots. Recall that to use the grid.arrange function, the gridExtra pacakge needs to be loaded.

```{r}
#CHUNK 3
library(gridExtra)
p1 <- ggplot(data = data.mortality, aes(x = as.factor(year), y = issage, fill = as.factor(year))) +
        geom_boxplot(alpha = 0.5) 
p2 <- ggplot(data = data.mortality, aes(x = resind_ind, y = issage, fill = smoker)) +
        geom_boxplot(alpha = 0.5) 
p3 <- ggplot(data = data.mortality, aes(x = sex, y = issage, fill = sex)) +
        geom_boxplot()
p4 <- ggplot(data = data.mortality, aes(x = smoker, y = issage, fill = smoker)) +
        geom_boxplot()
grid.arrange(p1,p2,p3,p4,ncol=2)
```

Run CHUNK 4 to learn how some variables relate to the actual face amount paid.

```{r}
#CHUNK 4
p1 <- ggplot(data = subset(data.mortality, actual_cnt >= 1), aes(x = resind_ind, y = actual_face, fill = resind_ind)) +
        geom_boxplot() +
        scale_y_log10()
p2 <- ggplot(data = subset(data.mortality, actual_cnt >= 1), aes(x = uwkey, y = actual_face, fill = uwkey)) +
        geom_boxplot() +
        scale_y_log10() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
p3 <- ggplot(data = subset(data.mortality, actual_cnt >= 1), aes(x = sex, y = actual_face, fill = sex)) +
        geom_boxplot() +
        scale_y_log10()
p4 <- ggplot(data = subset(data.mortality, actual_cnt >= 1), aes(x = smoker, y = actual_face, fill = smoker)) +
        geom_boxplot() +
        scale_y_log10()
grid.arrange(p1,p2,p3,p4,ncol=2)
```

Run CHUNK 5 to produce both stacked and split histograms of issue age for each product category.

```{r}
#CHUNK 5
## Stacked histogram
p1 <- ggplot(data = data.mortality, aes(issage, fill=prodcat, ..density..)) +
       geom_histogram(binwidth = 5, alpha = 0.5)

## Split histogram
p2 <- ggplot(data = data.mortality, aes(issage, fill=prodcat, ..density..)) +
       geom_histogram(binwidth = 5, alpha = 0.5, position = "dodge")
grid.arrange(p1,p2,ncol=1)
```

Run CHUNK 6 to try this for issue state.

```{r}
#CHUNK 6
## Stacked histogram
p1 <- ggplot(data = data.mortality, aes(issage, fill=issstate, ..density..)) +
       geom_histogram(binwidth = 5, alpha = 0.5)

## Split histogram
p2 <- ggplot(data = data.mortality, aes(issage, fill=issstate, ..density..)) +
       geom_histogram(binwidth = 5, alpha = 0.5, position = "dodge")
p1
p2
```

Run CHUNK 7 to try this for smoking status.

```{r}
#CHUNK 7
## Stacked histogram
p1 <- ggplot(data = data.mortality, aes(issage, fill=smoker, ..density..)) +
       geom_histogram(binwidth = 5, alpha = 0.5)

## Split histogram
p2 <- ggplot(data = data.mortality, aes(issage, fill=smoker, ..density..)) +
       geom_histogram(binwidth = 5, alpha = 0.5, position = "dodge")
grid.arrange(p1,p2,ncol=1)
```

Run CHUNK 8 to see a stacked bar chart.

```{r}
#CHUNK 8
p1 <- ggplot(data = data.mortality, aes(x = prodcat, fill = smoker)) +
        geom_bar(stat = "count", alpha = 0.5)
p2 <- ggplot(data = data.mortality, aes(x = prodcat, fill = smoker)) +
        geom_bar(stat = "count", alpha = 0.5, position = "dodge")
p3 <- ggplot(data = data.mortality, aes(x = uwtype, fill = uwkey)) +
        geom_bar(stat = "count", alpha = 0.5)
p4 <- ggplot(data = data.mortality, aes(x = uwtype, fill = uwkey)) +
        geom_bar(stat = "count", alpha = 0.5, position = "dodge")
grid.arrange(p1,p2,ncol=2)
grid.arrange(p3,p4,ncol=2)
```

Run CHUNK 9 to see an example of a heat map of issue age and duration.

```{r}
#CHUNK 9
# Calculate the kernel density of the two variables issue age and duration
library(MASS)
library(reshape2)
kernalDensity <- kde2d(data.mortality$issage, data.mortality$duration, n = 50)

# The following code turns the data from the k2de2d function into something that can be plotted by ggplot2
density.melt <- melt(kernalDensity$z)
density.melt$issage <- sapply(density.melt$Var1, FUN = function(a){kernalDensity$x[a]}) 
density.melt$duration <- sapply(density.melt$Var2, FUN = function(a){kernalDensity$y[a]})

# Plot the heat map
ggplot(data = density.melt, aes(x = issage, y = duration, fill = value)) +
        geom_tile() + 
        theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

Run CHUNK 10 to make a scatterplot.

```{r}
#CHUNK 10
p1 <- ggplot(data = data.mortality, aes(x = issage, y = duration)) +
        geom_point(alpha = 0.2)
p1
```

Run CHUNK 11 to do the same for duration and actual face amount.

```{r}
#CHUNK 11
p1 <- ggplot(data = data.mortality, aes(x = duration, y = actual_face)) +
        geom_point(alpha = 0.2) +
        scale_y_log10()
p1
```

Run CHUNK 12 to make two plots, one splitting by smoker, one by sex.

```{r}
#CHUNK 12
p1 <- ggplot(data = data.mortality, aes(x = duration, y = actual_face, color = smoker)) +
        geom_point(alpha = 0.2) +
        scale_y_log10()
p2 <- ggplot(data = data.mortality, aes(x = duration, y = actual_face, color = sex)) +
        geom_point(alpha = 0.2) +
        scale_y_log10()
grid.arrange(p1,p2,ncol=2)
```

Run CHUNK 13 to look at subsets (note the axis limits on the graphs).

```{r}
#CHUNK 13
p1 <- ggplot(data = subset(data.mortality, prodcat == "TRM"), aes(x = duration, y = actual_face, color = smoker)) +
        geom_point(alpha = 0.5) +
        scale_y_log10()
p2 <- ggplot(data = subset(data.mortality, prodcat == "TRM"), aes(x = duration, y = actual_face, color = sex)) +
        geom_point(alpha = 0.5) +
        scale_y_log10()
p3 <- ggplot(data = subset(data.mortality, prodcat == "TRM"), aes(x = duration, y = actual_face, color = smoker)) +
        geom_point(alpha = 0.5) +
        scale_x_continuous(limits = c(0, 80)) + 
        #scale_y_continuous(limits = c(0, 1.0e+07)) +
        scale_y_log10(limits = c(1, 1.0e+07))
p4 <- ggplot(data = subset(data.mortality, prodcat == "TRM"), aes(x = duration, y = actual_face, color = sex)) +
        geom_point(alpha = 0.5) +
        scale_x_continuous(limits = c(0, 80)) + 
        #scale_y_continuous(limits = c(0, 1.0e+07)) +
        scale_y_log10(limits = c(1, 1.0e+07))
grid.arrange(p1,p2,p3,p4,ncol=2)
```

Run CHUNK 14 to generate a heatmap of the correlations. 

```{r}
#CHUNK 14
# Correlations only work with numeric variables so we need to identify these first:
library(reshape2)
vars.numeric <- colnames(data.mortality)[sapply(data.mortality, is.numeric)]

# Calculate correlation matrix
cormat <- cor(data.mortality[vars.numeric])
# Melt the correlation matrix
melted.cormat <- melt(cormat)

# Plot the heat map
ggplot(data = melted.cormat, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
        hjust = 1))
```

Run CHUNK 15 to examine duration versus actual face for term insurance products.

```{r}
#CHUNK 15
p1 <- ggplot(data = subset(data.mortality, prodcat == "TRM"), aes(x = duration, y = actual_face)) +
        geom_point(alpha = 0.5) +
        scale_x_continuous(limits = c(0, 80)) + 
        scale_y_log10(limits = c(1, 1.0e+07)) +
        annotate("rect", xmin = 39, xmax = 43, ymin = 600, ymax = 10000, fill = "red", alpha = 0.2)
p1
```

Run CHUNK 16 to see an example where one variable is categorical and the other is numeric.

```{r}
#CHUNK 16
p1 <- ggplot(data = subset(data.mortality, actual_cnt >= 1), aes(x = smoker, y = issage, fill = smoker)) +
        geom_boxplot() 
p2 <- ggplot(data = subset(data.mortality, actual_cnt >= 1), aes(x = "smoker", y = issage)) +
        geom_boxplot() 
grid.arrange(p1,p2,ncol=2)
```

Run CHUNK 17 to make a frequency table of distribution channel and smoking status.

```{r}
#CHUNK 17
tbl <- table(data.mortality$distchan, data.mortality$smoker)
tbl
```

