---
title: 'Module 4.1: Univariate Data Exploration'
output:
  html_document: default
  html_notebook: default
---

Run CHUNK 1 to load and display Anscombe's Quartet.

```{r echo = FALSE}
#CHUNK 1
data.AQ <- read.csv("AnscombesQuartet.csv")
data.AQ
```

Run CHUNK 2 to determine the sample means and variances of each variable.

```{r echo = FALSE}
#CHUNK 2
data.frame(mean = sapply(data.AQ, mean),
           var = sapply(data.AQ, var))
```

Run CHUNK 3 to plot the variables

```{r echo = FALSE}
#CHUNK 3
library(ggplot2)
library(gridExtra)
p1 <- ggplot(data = data.AQ, aes(x = X1x, y = X1y)) +
       geom_point(color = 1) + ylim(2,14)
p2 <- ggplot(data = data.AQ, aes(x = X2x, y = X2y)) +
       geom_point(color = 2) + ylim(2,14)
p3 <- ggplot(data = data.AQ, aes(x = X3x, y = X3y)) +
       geom_point(color = 3) + ylim(2,14)
p4 <- ggplot(data = data.AQ, aes(x = X4x, y = X4y)) +
       geom_point(color = 4)+ ylim(2,14)
grid.arrange(p1,p2,p3,p4,ncol=2) 
```

Run CHUNK 4 to load the mortality data.

```{r echo = FALSE}
#CHUNK 4
data.mortality <- read.csv("soa_mortality_data.csv")
```

Run CHUNK 5 to see the output from the summary function.

```{r}
#CHUNK 5
summary(data.mortality$issage)
```

Run CHUNK 6 to calculate a variety of statistics.

```{r}
#CHUNK 6
stats.central <- data.frame(mean = mean(data.mortality$issage), 
                            median = median(data.mortality$issage),
                            percentile5 = quantile(data.mortality$issage, 0.05),
                            percentile25 = quantile(data.mortality$issage, 0.25),
                            percentile50 = quantile(data.mortality$issage, 0.50),
                            percentile75 = quantile(data.mortality$issage, 0.75),
                            percentile95 = quantile(data.mortality$issage, 0.95),
                            max = max(data.mortality$issage),
                            min = min(data.mortality$issage), 
                            row.names = c("issage"))
stats.central
```

Run CHUNK 7 to calculate some measures of spread.

```{r}
#CHUNK 7
library(e1071)
stats.spread <- data.frame(variance = var(data.mortality$issage),
                           standardDev = sd(data.mortality$issage),
                           iqr = IQR(data.mortality$issage),
                           skew = skewness(data.mortality$issage),
                           kurtosis = kurtosis(data.mortality$issage),
                           row.names = c("issage"))
stats.spread
```

Run CHUNK 8 to make a histogram of issue age.

```{r}
#CHUNK 8
p2 <- ggplot(data = data.mortality, aes(issage, ..density..)) +
       geom_histogram(binwidth = 5, fill = "blue", alpha = 0.5)
p2

```

Run CHUNK 9 to see different bin widths.

```{r}
#CHUNK 9
p1 <- ggplot(data = data.mortality, aes(issage, ..density..)) +
       geom_histogram(binwidth = 1, fill = "blue", alpha = 0.5)
p2 <- ggplot(data = data.mortality, aes(issage, ..density..)) +
       geom_histogram(binwidth = 20, fill = "blue", alpha = 0.5)
grid.arrange(p1,p2,ncol=2)

```

Run CHUNK 10 to make histograms of duration and attained age.

```{r}
#CHUNK 10
p1 <- ggplot(data = data.mortality, aes(duration, ..density..)) +
       geom_histogram(binwidth = 10, fill = "blue", alpha = 0.5)
p2 <- ggplot(data = data.mortality, aes(attage, ..density..)) +
       geom_histogram(binwidth = 10, fill = "blue", alpha = 0.5)
grid.arrange(p1,p2,ncol=2)

```

Run CHUNK 11 to split the variables by product category.


```{r}
#CHUNK 11
p1 <- ggplot(data = data.mortality, aes(x = prodcat, y = attage)) +
        geom_boxplot(fill = c(1:4), alpha = 0.5) 
p2 <- ggplot(data = data.mortality, aes(x = prodcat, y = issage)) +
        geom_boxplot(fill = c(1:4), alpha = 0.5) 
p3 <- ggplot(data = data.mortality, aes(x = prodcat, y = duration)) +
        geom_boxplot(fill = c(1:4), alpha = 0.5) 
p4 <- ggplot(data = data.mortality, aes(x = "attage", y = attage)) +
        geom_boxplot(alpha = 0.5) 
p5 <- ggplot(data = data.mortality, aes(x = "issage", y = issage)) +
        geom_boxplot(alpha = 0.5) 
p6 <- ggplot(data = data.mortality, aes(x = "duration", y = duration)) +
        geom_boxplot(alpha = 0.5) 

grid.arrange(p1,p4,p2,p5,p3,p6,ncol=2)

```

Run CHUNK 12 to make a frequency table.

```{r}
#CHUNK 12
table(data.mortality$prodcat)

table(data.mortality$prodcat) / nrow(data.mortality)
```

Run CHUNK 13 to use exposures.

```{r}
#CHUNK 13
# Exposure-weighted counts
tbl <- aggregate(data.mortality[,c("exposure_cnt","exposure_face")], by = list(ProductCategory = data.mortality$prodcat), FUN = sum) 

# Exposure-weighted Percentages
tbl$exposure_cnt_p <- tbl$exposure_cnt / sum(tbl$exposure_cnt)
tbl$exposure_face_p <- tbl$exposure_face / sum(tbl$exposure_face)

# Counts for comparison
tbl$count <- as.numeric(table(data.mortality$prodcat))
tbl$count_p <- tbl$count / sum(tbl$count)

tbl

```

Run CHUNK 14 to turn the frequency table into a bar chart.

```{r}
#CHUNK 14
library(dplyr)
# Count summary
p1 <- ggplot(data = data.mortality, aes(x = prodcat)) +
  geom_bar(stat = "count", fill = c(1:4), alpha = 0.5)

# Summary by exposure count variable
df <- data.mortality %>% group_by(prodcat) %>% summarize(exposure.count = sum(exposure_cnt))
p2 <- ggplot(data = df, aes(x = prodcat, y = exposure.count)) +
  geom_bar(stat = "summary", fill = c(1:4), alpha = 0.5)

# Summary by face amount exposure variable
df <- data.mortality %>% group_by(prodcat) %>% summarize(exposure.face = sum(exposure_face))
p3 <- ggplot(data = df, aes(x = prodcat, y = exposure.face)) +
  geom_bar(stat = "summary", fill = c(1:4), alpha = 0.5)

grid.arrange(p1,p2,p3,ncol=3)

```

Run CHUNK 15 to check for outliers in the duration variable.

```{r echo = FALSE}
#CHUNK 15
p1 <- ggplot(data = data.mortality, aes(duration, ..density..)) +
       geom_histogram(binwidth = 1, fill = "blue", alpha = 0.5)
p2 <- ggplot(data = data.mortality, aes(x = "duration", y = duration)) +
        geom_boxplot(fill = "blue", alpha = 0.5) 
grid.arrange(p1,p2,ncol=2)
```

In CHUNK 16 modify the code from CHUNK 15 to check for outliers in the duration variable after a log transformation.

```{r echo = FALSE}
#CHUNK 16
p1 <- ggplot(data = data.mortality, aes(duration, ..density..)) +
       geom_histogram(binwidth = 1, fill = "blue", alpha = 0.5) 
p2 <- ggplot(data = data.mortality, aes(x = "duration", y = duration)) +
        geom_boxplot(fill = "blue", alpha = 0.5) 
grid.arrange(p1,p2,ncol=2)
```

Run CHUNK 17 to see the effect of an outlier on a regression line.

```{r echo = FALSE}
#CHUNK 17
#The x-values are uniform(0,10) random numbers
#The y-values are the x-values plus a standard normal variable.
sample.x <- runif(10, min = 0, max = 10)
sample.y <- sample.x + rnorm(10, 0, 1)
#The "real" regression line should be rougly the line y=x.
line.real <- lm(sample.y~sample.x)
#Now change the tenth y-value to 50, an outlier
sample.y[10] = 50
#Redo the regression.
line.bias <- lm(sample.y~sample.x)
#Plot the two regression lines.
plot(sample.y~sample.x)
abline(line.real, col = "blue")
abline(line.bias, col = "red")
```

Run CHUNK 18 to revisit the duration variable on a log scale.

```{r echo = FALSE}
p1 <- ggplot(data = data.mortality, aes(x = "duration", y = duration)) +
        geom_boxplot(fill = "orange", alpha = 0.5) +
        scale_y_log10()
p2 <- ggplot(data = data.mortality, aes(x = prodcat, y = duration)) +
        geom_boxplot(fill = "orange", alpha = 0.5) +
        scale_y_log10()
grid.arrange(p1,p2,ncol=2)
```

CHUNK 19 makes a graph that illustrates the difference between squared and absolute error. It is provided here in case you are interested in how the graph was made.

```{r echo = FALSE}
x <- seq(-1.5,1.5,0.1)
errors <- data.frame(error = x,
                     squared = x^2,
                     abs = abs(x))

ggplot(data = errors, aes(x = error)) +
  geom_line(aes(y = squared), color = "red") +
  geom_line(aes(y = abs), color = "blue") +
  geom_line(aes(y = 1)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = 1, fill = "yellow", alpha = 0.2) + 
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 1, ymax = Inf, fill = "orange", alpha = 0.2)

```

