---
title: "Module 3 Practice"
output:
  html_notebook: default
  html_document: default
---

Run CHUNK 1 to load and save the MEPS data in a useful form. This can take some time, so skip this step if you did it in Section 3.5.
```{r}
#CHUNK 1
library(foreign)
meps.Data<-read.xport("../data/h171.ssp")
save(meps.Data,file="H171.Rdata")
```

Load the saved MEPS data using CHUNK 2 (if you ran CHUNK 1 it is already loaded).

```{r}
#CHUNK 2
load("../data/H171.Rdata")
```

Run CHUNK 3 to reduce the dataset to some interesting variables.
```{r}
library(caret)
# Count the rows and columns
nrow(meps.Data) # 34875 rows
ncol(meps.Data) # 1838 variables

# Hold out a sample of the data
set.seed(255)
holdout <- createDataPartition(meps.Data$TOTTCH14, p=0.3)
meps.Holdout <- meps.Data[ holdout$Resample1, ]
meps.Data <- meps.Data[ -holdout$Resample1, ]

nrow(meps.Holdout) # 10465 rows
nrow(meps.Data) # 24410 rows


# Some interesting variables include:
# PID - Person number
# SEX
# FAMINC14 - Family income
# TTLP14X - Persons income
# AGE14X - Age as of 31Dec2014
# BMINDX53- Adult BMI
# 
# INDCAT31 - Industry group
# OCCCAT31 - Occupation group
# TOTTCH14 - Total charged across all categories
# IPDIS14 - Total Hospital Discharges
# IPNGTD14 - Total number of nights in the hospital
# DVTOT14 - Total Dental Visits

# Make smaller data set with simpler names
meps.Simple <- data.frame(meps.Data$PID)
meps.Simple$sex <- meps.Data$SEX
meps.Simple$familyIncome <- meps.Data$FAMINC14
meps.Simple$personIncome <- meps.Data$TTLP14X
meps.Simple$age <- meps.Data$AGE14X
meps.Simple$bmi.adult <- meps.Data$BMINDX53
meps.Simple$bmi.child <- meps.Data$CHBMIX42
meps.Simple$industry <- meps.Data$INDCAT31
meps.Simple$occupation <- meps.Data$OCCCAT31
meps.Simple$totalCharges <- meps.Data$TOTTCH14
meps.Simple$totalHospDischarges <- meps.Data$IPDIS14
meps.Simple$totalNightsCare <- meps.Data$IPNGTD14
head(meps.Simple)

```

Use CHUNK 4 to evaluate the distribution of Age.

```{r}
#CHUNK 4
# Evaluate the age variable
typeof(meps.Simple$age) # double

# Calculate the mean
mean(meps.Simple$age) # Mean age is 34.8 years

# Get a summary of age
summary(meps.Simple$age) # Minimum value is '-1'

# Create a logical vector to identify the -1 values for age
is.neg1 <- meps.Simple$age == -1
typeof(is.neg1) # logical

meps.Simple[is.neg1,]$age # Filter the data to include only the '-1' values

# Count the '-1' values
length(meps.Simple[is.neg1,]$age) # 149
length(meps.Simple[is.neg1,]$age)/nrow(meps.Simple) # 0.006 or 0.6%

# How many distinct ages are in the data set
length(unique(meps.Simple$age)) # 87 values including -1


# Look at age distribution
hist(meps.Simple$age)
```

Use CHUNK 5 to delete records for wich Age = -1.

```{r}
#CHUNK 5
meps.Simple <- meps.Simple[!is.neg1,]
nrow(meps.Simple)
```

Use CHUNK 6 to explore the other variables and make adjustments as appropriate. Note that the exploration of age is based on knowing it can only be non-negative whole numbers. Recall that if you put the cursor on a line of code and hit Control-r only that line will be executed.

```{r}
#CHUNK 6
#Name the variable to investigate
variable.name <- c("age")

#Idenify which coumn it is in
col.number <- match(variable.name,colnames(meps.Simple))

# Evaluate the variable
typeof(meps.Simple[,col.number])

# Get a summary of the variable
summary(meps.Simple[,col.number]) 

# Create a logical vector to identify the -1 (or other interesting) values 
is.neg1 <- meps.Simple[,col.number] < 0
typeof(is.neg1) 
length(meps.Simple[is.neg1,col.number]) #Filter and count the number equal to -1 

# How many distinct values are there?
length(unique(meps.Simple[,col.number])) 

# Look at variable's distribution
hist(meps.Simple[,col.number])

#delete rows where the earlier condition is met
meps.Simple <- meps.Simple[!is.neg1,]
nrow(meps.Simple)
```

Use CHUNK 7 to create a single bmi value and change the sex values from 1 and 2 to M and F. 
If your dataset has a different name from meps.Simple, change the name by doing newname <- oldname and then rm(oldname).

```{r}
#CHUNK 7
library(dplyr)
# Create a single bmi variable from the adult and child bmi variables
meps.Simple[meps.Simple$bmi.adult <= 0, 'bmi.adult'] <- NA
meps.Simple$bmi <- coalesce(meps.Simple$bmi.adult, meps.Simple$bmi.child)

# Change the coding for sex, then summarize and make a simple plot

meps.Simple$sex[meps.Simple$sex == 1] <- 'M'
meps.Simple$sex[meps.Simple$sex == 2] <- 'F'
meps.Simple$sex <- as.factor(meps.Simple$sex)

summary(meps.Simple$sex)
plot(meps.Simple$sex)

# This is the last of the transformations we will do to this data during this practice so we will save the data file here
save(meps.Simple,file="my_meps.Simple.Rdata")

```

Use CHUNK 8 to create a variety of plots

```{r}
#CHUNK 8

# Load the pre-prepared data set provided through the link in the e-learning slides
load("../data/meps.Simple.Rdata")

library(ggplot2)

# Plot ages for each sex
ggplot(meps.Simple, aes(x=age, fill=sex)) + geom_histogram()
ggplot(meps.Simple, aes(x=age, fill=sex)) + geom_density(alpha = 0.2)

# Plot BMI
ggplot(meps.Simple[meps.Simple$bmi > 0,], aes(x=bmi)) + geom_histogram(binwidth = 1)

# Plot bmi with respect to age
ggplot(meps.Simple, aes(x=age, y=bmi)) + geom_point()

# This is an example of overplotting.  The points overlap and we cannot tell if there are 1 or 100 points at any particular area of the plot

# We could plot make some points transparent
ggplot(meps.Simple, aes(x=age, y=bmi)) + geom_point(alpha = 0.2)

# We can also create a heatmap that quantifies the points in each grid of the plot similar to a histogram
ggplot(meps.Simple, aes(x=age, y=bmi)) + geom_bin2d()
```

Run CHUNK 9 to see some vizualizations of the target variable.

```{r}
# Summarize total charges
summary(meps.Simple$totalCharges) # Note the wide range 0 to 1,990,000

# Plot total charges
ggplot(meps.Simple, aes(x=totalCharges)) + geom_histogram() # not a very informative plot

# Count the number of records with 0 totalCharges
nrow(meps.Simple[meps.Simple$totalCharges == 0, ]) # 5342 records
nrow(meps.Simple[meps.Simple$totalCharges == 0, ])/nrow(meps.Simple) # 0.2202 or 22.02%

# Let us transform the data using log and square root
summary(log(meps.Simple$totalCharges)) # The range of the log transformed value is more reasonable if we ignore those -Inf
summary(sqrt(meps.Simple$totalCharges))

ggplot(meps.Simple, aes(x=log(totalCharges))) + geom_histogram(bins = 200)

ggplot(meps.Simple, aes(x=sqrt(totalCharges))) + geom_histogram(bins = 200) # So many zeros

# 2 dimensional bins to deal with overplotting
ggplot(meps.Simple, aes(x=bmi, y=log(totalCharges))) + geom_bin2d(bins = 100)
ggplot(meps.Simple, aes(x=bmi, y=sqrt(totalCharges))) + geom_bin2d(bins = 100)

# Plot total charges with respect to age
 ggplot(meps.Simple[meps.Simple$bmi > 0,], aes(x=age, y=log(totalCharges))) + geom_point(alpha = 0.2)
```
