---
title: "Module 3.6 Data Design: Time Frame, Sampling and Granularity"
output: html_notebook
---

Run CHUNK 1 to load the gapminder dataset used previously. It should be in your working directory.

```{r}
#CHUNK 1
# Data
#data.gmd <- read.csv("gapminderDataFiveYear.csv")
data.gmd <- read.csv("gapminderDataFiveYear.csv")
```

Each of CHUNKS 2-6 generates a sample from this dataset. Try and identify from the code what type of sample is generated in each case.

```{r}
#CHUNK 2
library(caret)
data.split <- split(data.gmd, data.gmd$continent)
samples <- lapply(data.split, function(x) x[sample(1:nrow(x), size = 5, replace = FALSE),])
data.sample1 <- do.call(rbind,samples)
data.sample1
```

```{r}
#CHUNK 3
data.sample2 <- data.gmd[sample(1:nrow(data.gmd), size = 20, replace = FALSE),]
data.sample2
```

```{r}
#CHUNK 4
data.sample3 <- data.gmd[data.gmd$continent == "Africa",]
data.sample3
```

```{r}
#CHUNK 5
data.sub1 <- data.gmd[data.gmd$continent != "Oceania",]
data.sub2 <- data.gmd[data.gmd$continent == "Oceania",]

data.sample4 <- rbind(data.sub1[sample(1:nrow(data.sub1), size = 1000, replace = FALSE),],
                      data.sub2[sample(1:nrow(data.sub2), size = 1000, replace = TRUE),])
data.sample4
```

```{r}
#CHUNK 6
library(caret)
samples <- createDataPartition(data.gmd$continent, p = 0.1)

data.sample5 <- data.gmd[ samples$Resample1, ]
data.sample5
```

run CHUNK 7 to combine the two datasets.

```{r echo = FALSE}
#CHUNK 7
data.hospital <- read.csv("Patient_survey_full.csv")
data.stateMapping <- read.csv("state_table.csv")
data.hospital <- data.hospital[data.hospital$HCAHPS.Question == "Summary star rating" &
                                 data.hospital$Patient.Survey.Star.Rating != "Not Available",]
data.hospital$Patient.Survey.Star.Rating <- as.numeric(data.hospital$Patient.Survey.Star.Rating)
data.hospital <- merge(x = data.hospital, y = data.stateMapping, by.x = "State", by.y = "abbreviation", all.x = TRUE, all.y = FALSE)
```

Run CHUNK 8 to generate the graph by Zip Code.

```{r echo = FALSE}
#CHUNK 8
library(gridExtra)
library(dplyr)
df <- data.hospital %>% group_by(ZIP.Code) %>% summarize(Patient.Survey.Star.Rating = mean(Patient.Survey.Star.Rating))
# The above code calculates the average rating for each different value of zip code.

g1 <- ggplot(data = df, aes(x = ZIP.Code, y = Patient.Survey.Star.Rating)) +
  geom_point() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  ggtitle("Graph 1 - Rating")

g2 <- ggplot(data = data.hospital, aes(x = ZIP.Code)) +
  geom_bar(stat = "count", color = "blue", alpha = 0.5) +
  theme(axis.title.x=element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
  ggtitle("Graph 1 - Number of records")

grid.arrange(g1,g2,ncol=2)



```

Run CHUNK 9 to bring in the median income data for each zip code.

```{r echo = FALSE}
#CHUNK 9
data.income <- read.csv("MedianZIP-3.csv")
data.hospital <- merge(x = data.hospital, y = data.income, by.x = "ZIP.Code", by.y = "Zip", all.x = TRUE, all.y = FALSE)
data.hospital <- data.hospital[!(is.na(data.hospital$Median)),]
```

Run CHUNKS 10-13 to see graphs based on four levels of granularity.

```{r echo = FALSE}
#CHUNK 10
# Zip Code

df <- data.hospital %>% group_by(Median) %>% summarize(AverageRating = mean(Patient.Survey.Star.Rating))
ggplot(data = df[order(df$Median),], aes(x = Median, y = AverageRating)) + geom_line() +
  ggtitle("Zip Code")

```

```{r echo = FALSE}
#CHUNK 11
#County

df <- data.hospital %>% group_by(County.Name) %>% summarize(Median = mean(Median), AverageRating = mean(Patient.Survey.Star.Rating))
df <- df[!is.na(df$County.Name), ]
ggplot(data = df[order(df$Median),], aes(x = Median, y = AverageRating)) + geom_line() +
  ggtitle("County")

```

```{r echo = FALSE}
#CHUNK 12
# State

df <- data.hospital %>% group_by(State) %>% summarize(Median = mean(Median), AverageRating = mean(Patient.Survey.Star.Rating))
df <- df[!is.na(df$State), ]
ggplot(data = df[order(df$Median),], aes(x = Median, y = AverageRating)) + geom_line() +
  ggtitle("State")

```

```{r echo = FALSE}
#CHUNK 13
# Region

df <- data.hospital %>% group_by(census_region) %>% summarize(Median = mean(Median), AverageRating = mean(Patient.Survey.Star.Rating))
df <- df[!is.na(df$census_region), ]
ggplot(data = df[order(df$Median),], aes(x = Median, y = AverageRating)) + geom_line() +
  ggtitle("Region")

temp <- aggregate(data.hospital[, c("Median", "Patient.Survey.Star.Rating")], by = list(region = data.hospital$census_region), FUN = mean)
ggplot(data = temp[order(temp$Median),], aes(x = Median, y = Patient.Survey.Star.Rating)) +
  geom_line(stat = "summary", fun.y = mean) +
  ggtitle("Region")
```
