.lib---
title: "Module 3.8 Data Visualization"
output: html_notebook
---
Run CHUNK 1 to look at two datasets.

```{r Load datasets and ggplot2}
#CHUNK 1
library(datasets)
library(ggplot2)

# Display "R data sets" in a new tab
data()
```

Run CHUNK 2 to open and view a dataset from "datasets."

```{r Open and view iris from datasets}
#CHUNK 2
iris <- datasets::iris
head(iris)
```

Run CHUNK 3 to open and view a dataset from "ggplot2."

```{r Open and view economics from ggplot2}
#CHUNK 3
economics <- ggplot2::faithfuld
head(faithfuld)
```

You don't really need to load and assign and view these embedded datasets, as they can be used in functions directly. Run CHUNK 9 for an example:
  
```{r Plot AirPassengers}
#CHUNK 4
plot(AirPassengers)
```
 
Nonetheless, it is still good practice to do so, as you will be mostly working with data outside of these samples in real life.

In CHUNK 5 you should set the working director where you want it to be.
  
```{r Exercise - working directory}
#CHUNK 5
# Get the working directory 
  # Change it to the right location if needed to


```  

CHUNK 6 has you write code to learn some key facts about the diamonds dataset. You may have to search for help to get the right commands.

```{r Exercise - load and inspect data}
#CHUNK 6
# Load diamonds data and assign it to 'diamonds'


# How many rows and columns are in the data?


# Display the variable names and few sample records


# What do these variables mean?


```

CHUNK 7 has a model solution.

```{r Answer - load the data}
#CHUNK 7
# Get the working directory 
  # Change it to the right location if needed to
getwd()
#setwd("~/Desktop/SOA")

# Load the diamonds data
diamonds <- as.data.frame(diamonds)

# How many rows and columns are in the data?
dim(diamonds)

# Display the variable names in this data
head(diamonds)

# Help!
help("diamonds")
```

Be sure to note where the various output items appear.

Run CHUNK 8 to make a scatterplot.

```{r ggplot scatter plot}
#CHUNK 8
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point()
```

Run CHUNK 9 to add color to the previous graph.

```{r ggplot scatter plot - add color}
#CHUNK 9
ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) +
  geom_point()
```
  
Run CHUNK 10 to map color to shape. Try writing your own code before looking at a solution in CHUNK 11.

```{r Exercise - use shape to differentiate cut}
#CHUNK 10


```




```{r Answer - use shape to differentiate cut}
#CHUNK 11
ggplot(data = diamonds, aes(x = carat, y = price, shape = cut)) +
  geom_point()
```

Write the code in CHUNK 12 to differentiate cut by using size. Of color, shape, and size, which is most effective?

```{r Exercise - use size to differentiate cut}
#CHUNK 12

```

Run CHUNK 13 to see the difference between items specified within aes() and those outside.

```{r ggplot scatter plot - change the dots to color red}
#CHUNK 13
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(color = 'red') # Note color is not within aes()
```

Run CHUNK 14 to see the addition of a line.

```{r ggplot scatter plot - add a line}
#CHUNK 14
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point() +
  geom_smooth() 
```

CHUNK 15 provides an execise for you to try some other options.

```{r Exercise - use geom_line instead of geom_smooth}
#CHUNK 15
# Use help() if you don't understand the differences

```

CHUNK 16 has a solution.

```{r Answer - use geom_line instead of geom_smooth}
#CHUNK 16
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point() +
  geom_line() 
# This simply connects the dots, instead of smoothing it. 
# You can add stats functions within geom_line() to display plots similar to those in geom_smooth (),
# and you will see situations where both approaches are useful.
# Just remember to always keep your code clean, short and simple. 
```
  
Run CHUNK 17 to change the transparency of the points.  
  
```{r ggplot scatter plot - alpha}
#CHUNK 17
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 1/30)
```

Run CHUNK 18 to add jitter.
  
```{r ggplot scatter plot - jitter}
#CHUNK 18
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 1/30, position = position_jitter(width = 0.2, height = 0))
```

Run CHUNK 19 to implement facet_wrap on cut.

```{r ggplot scatter plot - facet_wrap}
#CHUNK 19 
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 1/30) +
  geom_smooth() +
  facet_wrap(~cut)
```

Use facet_wrap on another variable and share your observations afterward. CHUNK 20 is set up for you to do that.

```{r Answer - facet_wrap}
#CHUNK 20
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 1/30) +
  geom_smooth() +
  facet_wrap()
```

Run CHUNKS 21-23 to see three examples of facet_grid.

```{r ggplot scatter plot - facet_grid1}
#CHUNK 21
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 1/30) +
  facet_grid(.~cut)
# You can see that the Y-axis is common for all cut comparisons, this is not necessarily true for facet_wrap. 
# The dot in the parenthesis indicates no faceting on this axis. 
```

```{r ggplot scatter plot - facet_grid2}
#CHUNK 22
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 1/30) +
  facet_grid(color~.)
# You can see the X-axis is common for all color comparisons, this is not necessarily true for facet_wrap.
# This time, the dot indicates no faceting on x-axis.
```

```{r ggplot scatter plot - facet_grid3}
#CHUNK 23
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 1/30) +
  facet_grid(color~cut)
# Faceting both x and y axes. 
```

Use CHUNK 24 to try and add features to the graph. A solution is in CHUNK 25.
  
```{r Exercise - add title, labels and legends}
#CHUNK 24
ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = 1/30) +
  geom_smooth()

```


```{r Sample Answer - add title, labels and legends}
#CHUNK 25
ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = 1/30) +
  geom_smooth() +
  ggtitle("Diamonds are expensive!") + 
  xlab("Weight in Carats") +           
  ylab("Price in USD") +               
  scale_y_continuous(label = function(x){return(paste("The diamond is $", x))})
```
    
CHUNK 26 restricts the plot to diamonds with carats <= 2.5.

```{r Answer - carats less than or equal to 2.5}
#CHUNK 26
ggplot(data = subset(diamonds, carat <= 2.5), aes(x = carat, y = price)) +
  geom_point(alpha = 1/30)
```

Try adding xlim() to CHUNK 27 to produce the same result (including only those diamonds with carats <= 2.5).

```{r Exercise - use xlim}
#CHUNK 27
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 1/30)
```
  
CHUNK 28 provides a speed test of two ways to limit the graph to diamonds with carats <= 0.5.

```{r Sample Answer - 3.8.9}
#CHUNK 28
library(microbenchmark)
set.seed(1000)
microbenchmark(
p1 = { ggplot(data = subset(diamonds, carat <= 0.5), aes(x = carat, y = price)) +
         geom_point()},
p2 = { ggplot(data = diamonds, aes(x = carat, y = price)) +
         geom_point() + xlim(0, 0.5)},
times = 1000
)
```

Modify CHUNK 29 to cut out the highest 5% on both carat and price.

```{r Exercise - limit the axes using quantile}
#CHUNK 29
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 1/30)
```

CHUNK 30 has a solution.

```{r Answer - limit the axes using quantile}
#CHUNK 30
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 1/30) +
  xlim (c(0, quantile(diamonds$carat, probs = 0.95))) +
  ylim (c(0, quantile(diamonds$price, probs = 0.95)))
```

CHUNK 31 brings back an earlier version of the graph. Try to modify it to to plot y on a log scale.
 
```{r Exercise - log(price)}
#CHUNK 31
ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 1/30) +
  xlim(0, 2.5)
```

Run CHUNK 32 to see two ways to implement a log transform.
  
```{r Answer - log(price)}
#CHUNK 32
library(gridExtra)
p1 <- ggplot(data = diamonds, aes(x = carat, y = log(price))) +
        geom_point(alpha = 1/30) +
        xlim(0, 2.5)
p2 <- ggplot(data = diamonds, aes(x = carat, y = price)) +
        geom_point(alpha = 1/30) +
        xlim(0, 2.5) +
        scale_y_log10()
grid.arrange(p1, p2, ncol = 2)
```

Run CHUNK 33 to see an example of adding means to the scatterplot.

```{r ggplot scatter plot - add summary (mean)}
#CHUNK 33
library(dplyr)
df <- diamonds %>% group_by(carat) %>% summarize(meanprice = mean(price))
# The above code calculates the average price for each different value of carat.

ggplot(data = diamonds, aes(x = carat)) +
  geom_point(aes(y = price), alpha = 1/30) +
  xlim(0, 2.5) +  geom_line(data = df, aes(y = meanprice), color = "red")
# Note the use of two different data frames.
```

Modify CHUNK 34 to add the median and 25th and 75th percentiles.

```{r Exercise - add median, 25% and 75%}
#CHUNK 34
df <- diamonds %>% group_by(carat) %>% summarize(meanprice = mean(price))

ggplot(data = diamonds, aes(x = carat)) +
  geom_point(aes(y = price), alpha = 1/30) +
  xlim(0, 2.5) +  geom_line(data = df, aes(y = meanprice), color = "red")
```
  
CHUNK 35 provides a solution.

```{r Answer - add median, 25% and 75%}
#CHUNK 35
# Note we plot it this way to help you to see:
  # - the differences between mean and median
  # - the range of 25% and 75% increase for larger diamonds

df <- diamonds %>% group_by(carat) %>% summarize(meanprice = mean(price),
                                                 medianprice = median(price),
                                                 q1price = quantile(price, 0.25),
                                                 q3price = quantile(price, 0.75))

ggplot(data = diamonds, aes(x = carat)) +
  geom_point(aes(y=price), alpha = 1/30) +
  xlim(0, 2.5) +  geom_line(data = df, aes(y=meanprice), color = "red") +
  geom_line(data = df, aes(y = medianprice), color = "blue") +
  geom_line(data = df, aes(y = q1price), color = "orange") +
  geom_line(data = df, aes(y = q3price), color = "orange")



```
 
Run CHUNK 36 to make a bar chart of cut.

```{r Answer - number of diamonds by cut}
#CHUNK 36
ggplot(data = diamonds, aes(cut)) +
  geom_bar()
```

Edit CHUNK 37 make a bar chart to show the average diamond price for each cut.

```{r Exercise - average price by cut}
#CHUNK 37
ggplot(data = , aes()) +
  geom_bar()
```

In CHUNK 38 you can get the averages to use as a check.

```{r gglot bar chart - Average price by cut}
#CHUNK 38
# Self-check: does the value match what is displayed in the bar chart?
by(diamonds$price, diamonds$cut, mean)
```
  
And CHUNK 39 provides the code to make the desired chart.

```{r Answer - average price by cut}
#CHUNK 39
# Why does "Ideal" (in theory, the best cut) have the lowest average price?
# This gets us into uni-variate vs multi-variate analysis, which we will cover in later modules. 
# Try a few other variables as well, e.g. color, clarity.
ggplot(data = diamonds, aes(x = cut, y = price)) +
  geom_bar(stat = "summary")
```

Use CHUNK 40 to put the two graphs already made side-by-side.
  
```{r Exercise - two bar charts side by side}
#CHUNK 40
p1 <-
p2 <-
grid.arrange()

```

Run CHUNK 41 to make a grouped bar chart. Note the observations in the comments.

```{r Answer - average price by color grouped by cut}
#CHUNK 41
# Did you notice that the average price for each color is increasing as expected for ideal and premium cut, but not for the other cuts? 
# How about that the "Ideal" cut has the lowest average price in almost all color categories? 
ggplot(data = diamonds, aes(x = color, y = price, fill = cut)) +
  geom_bar(stat = "summary", position = position_dodge())

# Maybe it is because "Ideal" diamonds are on average smaller...plot another bar chart to confirm.
# Confirmed. But does that explain all the differences?
# This brings us to exploratory data analysis (EDA), which will be covered in Module 4. 
ggplot(data = diamonds, aes(x = cut, y = carat)) +
  geom_bar(stat = "summary")
```
  
For the same grouped bar chart, average price by color, grouped by cut and stack the bars instead of having them side by side. Also, stack them to 100% of average price. Try it out in CHUNK 43 and then look at CHUNK 44 for a solution and some comments.

```{r Exercise - stack 100% average price by color grouped by cut}
#CHUNK 42

```

  
```{r Answer - - stack 100% average price by color group by cut}
#CHUNK 43
# You probably noticed that this is not a good visualization to show "average price per color is increasing as expected for ideal and premium cut, but not for others".
# In this case, a facet_wrap works better. 
# You can alway try various visualizations either to observe the variables from different angles, or demonstrate an idea more clearly. 
ggplot(data = diamonds, aes(x = color, y = price, fill = cut)) +
  geom_bar(stat = "summary", position = "fill")

ggplot(data = diamonds, aes(x = color, y = price)) +
  geom_bar(stat = "summary") +
  facet_wrap(~cut)
```

Plot a histogram of price in CHUNK 44.

```{r Answer - diamond price histogram}
#CHUNK 44
#Note the use of ..density.., it ensures that a proper histogram is made.
ggplot(data = diamonds, aes(price, ..density..)) +
  geom_histogram()
```

Enhance CHUNK 45 to have plot 2 use 100 bins, plot 3 have bin boundaries running from 0 to 14,500 by 500 and then 15,000 to 20,000 by 1000, and plot 4 modify plot 1 to only show prices from 0 to 2000. 

```{r Sample Answer - histogram with different binning}
#CHUNK 45
# None of the sample answers are particularly good...
# We can see that, in the first graph, there's a bar missing
# p4 was added to check what was going on - it seems like there's no diamond priced at $1500
p0 <- ggplot(data = diamonds, aes(price, ..density..))
p1 <- p0 + geom_histogram(binwidth = 100)
p2 <- p0 + geom_histogram()
p3 <- p0 + geom_histogram()
p4 <- p1 + xlim(0, 2000)
grid.arrange(p1, p2, p3, p4, ncol = 2)
```

Run CHUNK 46 to apply a log transformation to price.

```{r Answer - log-price histogram}
#CHUNK 46
# It doesn't look so normal does it? 
# It could be that the diamond market is really split by two separate consumer groups, logical vs crazy ones!!!
# This type of observation has implications for how we might approach predicting the diamond price and so is a good example of how early visualisations can help mold subsequent analytics approaches.
# This is discussed in more detail in later modules. 
ggplot(data = diamonds, aes(price, ..density..)) +
  geom_histogram(bins = 200) +
  scale_x_log10()
```
  
 Create a 2D binned histogram in CHUNK 47.

```{r Sample Answer - 2D Histogram}
#CHUNK 47
# You can see that this plot is convenient for dealing with overlapping points
ggplot(data = diamonds, aes(x=carat, y=price)) + geom_bin2d()

```

Add a frequency plot to the original and transformed price histograms. Display them side by side. Try it in CHUNK 48 and see a solution in CHUNK 49.

```{r Exercise - frequency plot}
#CHUNK 48


```

```{r Answer - frequency plot}
#CHUNK 49
p0 <- ggplot(data = diamonds, aes(price, ..density..)) +
        geom_histogram(bins = 100)
p1 <- p0 + geom_freqpoly(bins = 100, color = "red", size = 2)
p2 <- p1 + scale_x_log10()
grid.arrange(p1, p2, ncol = 2)
```

Use the next two chuncks to change these to density plots.

```{r Exercise - density plot}
#CHUNK 50


```

```{r Answer - density plot}
#CHUNK 51
# The default kernel is Gaussian. Feel free to explore other kernels
# Why do you think there is a line on across the bottom of the plot?
p0 <- ggplot(data = diamonds, aes(price, ..density..)) +
        geom_histogram(aes(y = ..density..), bins = 100)
p1 <- p0 + geom_density(color = "red", size = 2)
p2 <- p1 + scale_x_log10()
grid.arrange(p1, p2, ncol = 2)
```
  
Run CHUNK 52 to make a box plot.

```{r box plot}
#CHUNK 52
set.seed(1000)
y=rnorm(1000)
norm.sim <- as.data.frame(y)
ggplot(data=norm.sim,aes(x="",y=y)) +
  geom_boxplot()
```

Use the blank area in CHUNK 53 to make one or two boxplots that display the variable price by color of the diamond.

```{r Exercise - box plot}
#CHUNK 53

```

Here is a solution.

```{r Answer - box plot}
#CHUNK 54
# As there are many "outliers" in the price plot, the transformed version looks better. 
# This confirms some of our previous observations
# - Median price increases as color upgrades
# - Distribution of color J is tighter than others
# - The better colors (H, I, J) exhibit more left skew after transformation
p1 <- ggplot(data = diamonds, aes(x = color, y = price)) +
        geom_boxplot()
p2 <- p1 + scale_y_log10()
grid.arrange(p1, p2, ncol = 2)
```

Run CHUNK 55 to see an exmaple of ggpairs.

```{r ggplot box plot - ggpairs}
#CHUNK 55
library(GGally)
ggpairs(diamonds, c(1:2, 7))
```

CHUNK 56 is a rerun from Module 1.1. The commands may be helpful as you try to create a similar plot for the diamonds data set. 
  
```{r Example from module 1.1}
#CHUNK 56
data.all <- read.csv("../data/BreastCancerWisconsinDataSet.csv")
 # set the target 
data.all$target [data.all$diagnosis == "M"] = 1
data.all$target [data.all$diagnosis == "B"] = 0

 # all variables available for training
vars <- names(data.all)[c(-1, -2, -33)]
data.all <- data.all[c(-1, -2, -33)]

# Calculate correlation matrix
library(reshape2)
cormat <- cor(data.all)
 # Melt the correlation matrix
melted.cormat <- melt(cormat)

ggplot(data = melted.cormat, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
        hjust = 1))
```
  
Now in CHUNK 57 try and create a similar plot for the diamond data.   

```{r Exercise - heat map}
#CHUNK 57

 
```

CHUNK 58 has one approach to creating numerical variables.

```{r Sample answer - change categorical variables to indicators}
#CHUNK 58

library(caret)
# dummyVars is not compatible with factors
diamonds$color <- as.character(diamonds$color)
diamonds$cut <- as.character(diamonds$cut)
diamonds$clarity <- as.character(diamonds$clarity)
# Binarize the variables
binarizer <- caret::dummyVars('~ cut + clarity + color', data = diamonds, fullRank= F)
binarized.vars <- data.frame(predict(binarizer, newdata = diamonds))
# Combine the binarized variables and the original data set
drop.vars <- c('cut', 'clarity', 'color')
diamonds_spread <- cbind(diamonds[ , !(names(diamonds)) %in% drop.vars], binarized.vars)

```

CHUNK 59 has an alternative approach.

```{r Sample answer - change categorical variables to numeric}
#CHUNK 59
diamonds_mod <- diamonds[c(-2, -3, -4)]
diamonds_mod$cut <- as.numeric(diamonds$cut)
diamonds_mod$color <- as.numeric(diamonds$color)
diamonds_mod$clarity <- as.numeric(diamonds$clarity)
```

CHUNKS 60 and 61 make the heat maps.

```{r Sample answer - heat map with diamonds_spread}
#CHUNK 60
# Calculate correlation matrix
cormat <- cor(diamonds_spread)
 # Melt the correlation matrix
melted.cormat.spread <- melt(cormat)
ggplot(data = melted.cormat.spread, aes(x=Var1, y=Var2, fill=value)) + 
    geom_tile() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

```{r Sample answer - heat map with diamonds_mod}
 #CHUNK 61
# Calculate correlation matrix
cormat <- cor(diamonds_mod)
 # Melt the correlation matrix
melted.cormat.mod <- melt(cormat)
ggplot(data = melted.cormat.mod, aes(x=Var1, y=Var2, fill=value)) + 
    geom_tile() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

CHUNK 62 makes the heat map with ggplot.
  
```{r ggplot heat map}
#CHUNK 62
ggplot(melted.cormat.spread, aes(x = Var1, y = Var2, fill = value)) + 
  geom_tile()

ggplot(melted.cormat.mod, aes(x = Var1, y = Var2, fill = value)) + 
  geom_tile()
```