---
title: "Module 3.7 Data Design: Data Blending/Matching"
output: html_notebook
---

Run CHUNK 1 to create and display a small life insurance dataset.

```{r echo=FALSE}
#CHUNK 1
policy.data <- data.frame(PolicyNumber = c(1, 2, 3, 4, 5),
                          Name = c("Tina Adams", "Mary Davis", "Josh Griffin", "Amy Fox", "Harry Goldberg"),
                          Zip = c("98801", "50046", "49884", "49969", "99344"),
                          Sex = c("Female", "Female", "Male", "Female", "Male"),
                          IssueAge = c(32, 51, 50, 43, 39))
policy.data
```

Run CHUNK 2 to create and display claims data.

```{r echo = FALSE}
#CHUNK 2
claims.data <- data.frame(PolicyNumber = c(1, 3),
                          ClaimID = c(1, 2),
                          ClaimAmount = c(40000, 60000))

claims.data
```

Run CHUNK 3 to merge the two datasets.

```{r}
#CHUNK 3
merged.data <- merge(x = policy.data, y = claims.data, by = "PolicyNumber", all.x = TRUE, all.y = FALSE)
merged.data
```

Run CHUNK 4 to generate a dataset of median incomes.

```{r echo = FALSE}
#CHUNK 4
demographic.data <- data.frame(ZipCode = c("98801", "50046", "49884", "49969", "99344","98801", "50046", "49884", "49969", "99344"),
                               Sex = c("Male", "Male", "Male", "Male", "Male", "Female", "Female", "Female", "Female", "Female"),
                               MedianIncome = c(65000, 55000, 40000, 45000, 75000, 60000, 30000, 42000, 40000, 63000))

demographic.data
```

Run CHUNK 5 to bring in the income data.

```{r}
#CHUNK 5
merged.data <- merge(x = merged.data, y = demographic.data, by.x = c("Zip", "Sex"), by.y = c("ZipCode", "Sex"), all.x = TRUE, all.y = FALSE)

merged.data
```

CHUNK E371 is available for your work on Exercise 3.7.1.

```{r}
#CHUNK E371
library(readxl)
county <- read_excel("Food_Environment_Atlas.xls",sheet = 'Supplemental Data - County')
state <- read_excel("Food_Environment_Atlas.xls",sheet = 'Supplemental Data - State')
health <- read_excel("Food_Environment_Atlas.xls",sheet = 'HEALTH')
colnames(state)[1] <- 'State' #Rename the first column

```

Run CHUNK 6 to redo the demographic data with a different coding for Sex.

```{r echo = FALSE}
#CHUNK 6
demographic.data <- data.frame(ZipCode = c("98801", "50046", "49884", "49969", "99344","98801", "50046", "49884", "49969", "99344"),
                               Sex = c("M", "M", "M", "M", "M", "F", "F", "F", "F", "F"),
                               MedianIncome = c(65000, 55000, 40000, 45000, 75000, 60000, 30000, 42000, 40000, 63000))

demographic.data
```

Run CHUNK 7, which uses the same merge command as before.

```{r}
#CHUNK 7
merged.data <- merge(x = policy.data, y = demographic.data, by.x = c("Zip", "Sex"), by.y = c("ZipCode", "Sex"), all.x = TRUE, all.y = FALSE)

merged.data
```

Run CHUNK 8 to modify the coding in the deomgrpahic file and then run the merge again.

```{r}
#CHUNK 8
# Map M to Male and F to Female
demographic.data$Sex <- ifelse(demographic.data$Sex == "M", 
                               "Male",
                               "Female")

merged.data <- merge(x = policy.data, y = demographic.data, by.x = c("Zip", "Sex"), by.y = c("ZipCode", "Sex"), all.x = TRUE, all.y = FALSE)

merged.data
```

Run CHUNK 9 to generate three addresses that include zip code.

```{r echo = FALSE}
#CHUNK 9
address.data <- data.frame(Address = c("1 Fireman St. 49884","Apt 5, 6 New Ave, 99344", "21 Grey Road 50046 USA"))
address.data
```

Run CHUNK 10 to see how a zip code can be extracted from an address.

```{r}
#CHUNK 10
## Create the regular expression
zipRegex <- "[0-9]{5}"

## Search through a string using the regular expression
library(stringr)
testString <- "1 Fireman St. 49884"
str_extract(testString, zipRegex)
```

Run CHUNK 11 to merge addresses.

```{r}
#CHUNK 11
## Map the address to zip code
zipRegex <- "[0-9]{5}"
address.data$Zip <- str_extract(address.data$Address, zipRegex)

## Merge datasets using the zip variable
merge(x = address.data, y = demographic.data, by.x = "Zip", by.y = "ZipCode", all.x = TRUE, all.y = FALSE)
```

Run CHUNK 12 to re-establish the policyholder data and add a new dataset with marriage status.

```{r}
#CHUNK 12
# data from the beginning of the Module
policy.data <- data.frame(PolicyNumber = c(1, 2, 3, 4, 5),
                          Name = c("Tina Adams", "Mary Davis", "Josh Griffin", "Amy Fox", "Harry Goldberg"),
                          Zip = c("98801", "50046", "49884", "49969", "99344"),
                          Sex = c("Female", "Female", "Male", "Female", "Male"),
                          IssueAge = c(32, 51, 50, 43, 39))

# Match the following data to policy.data
surname.data <- data.frame(Name = c("J Griffin", "A Fox", "M Davis", "T Adams", "H Goldberg"),
                           MarriageStatus = c("Married", "Single", "Single", "Married", "Defacto"))
surname.data
```

Use the space in CHUNK 13 to merge the two files.

```{r}
#CHUNK 13


```

Run CHUNK 14 to see a slightly different version of the marriage status file where the names aren't always accurate.

```{r echo = FALSE}
#CHUNK 14
names.data <- data.frame(Name = c("J. Griffin", "Amy Anne Fox", "M C Davis", "Tina Adams", "Harry Gouldberg"),
                         MarriageStatus = c("Married", "Single", "Single", "Married", "Defacto"))
names.data
```

Run CHUNK 15 to start the process of matching the names in the names.data dataset to those in the policy.data data set.

```{r}
#CHUNK 15
library(stringdist)

#Set up the names datase
names.data <- data.frame(Name = c("J. Griffin", "Amy Anne Fox", "M C Davis", "Tina Adams", "Harry Gouldberg"),
                         MarriageStatus = c("Married", "Single", "Single", "Married", "Defacto"))

# Set up the matrix for storing the results
scores <- matrix(NA, ncol = length(names.data$Name), nrow = length(policy.data$Name))


for(i in 1:length(names.data$Name)) {
    for(j in 1:length(policy.data$Name)) { 
      # Calate the distance between policy.data name j and names.data name i using the Levenshtein Distance
      scores[j,i] <- stringdist(tolower(names.data$Name[i]),tolower(policy.data$Name[j]), method = "lv")      
    }  
}

# View the results:
scoresPretty <- data.frame(scores, row.names = policy.data$Name)
colnames(scoresPretty) <- names.data$Name
scoresPretty
```

Run CHUNKS 16 and 17 to perform the merge using the scores.

```{r}
#CHUNK 16
# Map to the *best* match 
policy.data$NameMatched <- names.data$Name[apply(scores, 1, which.min)] 
# See ?apply and ?which.min to understand this line of code

policy.data

```

```{r}
#CHUNK 17
# Merge to the names dataset
merge(x = policy.data, y = names.data, by.x = "NameMatched", by.y = "Name")
```

Run CHUNKA 18 and 19 to set up two new datasets.

```{r echo = FALSE}
#CHUNK 18
data.frame(Address1 = address.data$Address)
```

```{r echo = FALSE}
#CHUNK 19
address2.data <- data.frame(Address2 = c("1 Fireman Street 49884",
                                         "9 Fireman St. 49884",
                                         "6 New Ave, Flat 5, 99344",
                                         "Apt 10, 6 Knowl Ave, 99344",
                                         "21 Grey Rd 50046",
                                         "27 Red Road 50046 USA"))

address2.data

```

Run CHUNK 20 to calculate the distances between these two sets of addresses.

```{r}
#CHUNK 20
# Set up the matrix for storing the results
scores <- matrix(NA, ncol = length(address.data$Address), nrow = length(address2.data$Address2))


for(i in 1:length(address.data$Address)) {
    for(j in 1:length(address2.data$Address2)) { 
      # Calate the distance between policy.data name j and names.data name i using the Levenshtein Distance
      scores[j,i] <- stringdist(tolower(address.data$Address[i]),tolower(address2.data$Address2[j]), method = "lv")      
    }  
}

# Display the results:
scoresPretty <- data.frame(scores, row.names = address2.data$Address2)
colnames(scoresPretty) <- address.data$Address
scoresPretty
```

Run CHUNKS 21 and 22 to create life and health policyholder datasets.

```{r echo = FALSE}
#CHUNK 21
life.data <- data.frame(CustomerID = c("A143", "A234", "B512", "A956", "B128", "C876"),
                        Name = c("Anne G. Hammerday", "Ben Johnson", "carol H. Davidson", "Fred S. Jones", "Georgia Stynes", "Diane J. Stevens"),
                        Email = c("anne.hammerday@company.com", "bj23@email.com", "c.davidson@email.com", "monster_truck_guy123@mail.com", "georgiastynes@xyz.com", ""),
                        PolicyID = c("L123", "L234", "L345", "L456", "L567", "L678"))
life.data
```

```{r echo = FALSE}
#CHUNK 22
health.data <- data.frame(CustomerID = c("345823", "846712", "0393521"),
                        Name = c("Anne Gloria Hammerday", "Georgia Patty Stynes", "Daniel Mark Smithson"),
                        Email = c("anne.hammerday@company.com", "g.stynes@email.com", "dms123456@email.com"),
                        PolicyID = c("H987", "H876", "H765"))
health.data
```

Run CHUNK 23 to merge based on email address.

```{r}
#CHUNK 23
merge(x = life.data, y = health.data, by = "Email", all.x = TRUE, all.y = FALSE )
```

Run CHUNK 24 to try a fuzzy merge based on name.

```{r}
#CHUNK 24
scores <- matrix(NA, ncol = length(life.data$Name), nrow = length(health.data$Name))

for(i in 1:length(life.data$Name)) {
    for(j in 1:length(health.data$Name)) { 
      # Calate the distance between policy.data name j and names.data name i using the Levenshtein Distance
      scores[j,i] <- stringdist(tolower(life.data$Name[i]),tolower(health.data$Name[j]), method = "lv")      
    }  
}

scoresPretty <- data.frame(scores, row.names = health.data$Name)
colnames(scoresPretty) <- life.data$Name
scoresPretty

life.data$NameMatched <- health.data$Name[apply(scores, 2, which.min)]

merge(x = life.data, y = health.data, by.x = "NameMatched", by.y = "Name", all.x = TRUE, all.y = FALSE)

```

Run CHUNK 25 to use a threshold of 10.

```{r}
#CHUNK 25
life.data$NameMatched[apply(scores, 2, min) >= 10] <- NA

merge(x = life.data, y = health.data, by.x = "NameMatched", by.y = "Name", all.x = TRUE, all.y = FALSE)
```


