---
title: "Module 5.1 Feature Generation and Selection: Introduction"
output: html_notebook
---

Run CHUNK 1 to load some packages that will be used later in this section.

```{r echo = FALSE}
#CHUNK 1
# Required libraries
library(ggplot2)
library(gridExtra)
library(scatterplot3d)
```



```{r echo = FALSE}
#CHUNK 2
df <- data.frame(target = c(1.5,1,2.2,1.9), 
                 var = c(1,2.2,1,2.5), 
                 feature = c(0.8,0.5,1.2,0.9),
                 col = c("Data Point 1", "Data Point 2", "Data Point 3", "Data Point 4"))

p1 <- ggplot(data = df, aes(x = var, y = target, color = col)) +
  geom_point(size = 3) +
  scale_x_continuous(limits = c(0,3)) +
  scale_y_continuous(limits = c(0,3)) +
  ggtitle("Input space")


p2 <- ggplot(data = df, aes(x = feature, y = target, color = col)) +
  geom_point(size = 3) +
  scale_x_continuous(limits = c(0,1.5)) +
  scale_y_continuous(limits = c(0,3)) +
  ggtitle("Feature space")


grid.arrange(p1,p2,ncol = 2)



```

Run CHUNKS 3A and 3B to see the effect of increasing dimensions without increasing the number of data points.

```{r echo = FALSE}
#CHUNK 3A
df <- data.frame(x1 = runif(30), x2 = runif(30), x3 = runif(30))

p1 <- ggplot(data = df, aes(x = x1, y = 1)) +
  geom_point() +
  annotate("segment", x = 0, xend = 0, y = 0.5, yend = 1.5) +
  annotate("segment", x = 0.2, xend = 0.2, y = 0.5, yend = 1.5) +
  annotate("segment", x = 0.4, xend = 0.4, y = 0.5, yend = 1.5) +
  annotate("segment", x = 0.6, xend = 0.6, y = 0.5, yend = 1.5) +
  annotate("segment", x = 0.8, xend = 0.8, y = 0.5, yend = 1.5) +
  annotate("segment", x = 1, xend = 1, y = 0.5, yend = 1.5) +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  ggtitle("1D - 5 Regions")

p2 <- ggplot(data = df, aes(x = x1, y = x2)) +
  geom_point() +
  annotate("segment", x = 0, xend = 0, y = 0, yend = 1) +
  annotate("segment", x = 0.2, xend = 0.2, y = 0, yend = 1) +
  annotate("segment", x = 0.4, xend = 0.4, y = 0, yend = 1) +
  annotate("segment", x = 0.6, xend = 0.6, y = 0, yend = 1) +
  annotate("segment", x = 0.8, xend = 0.8, y = 0, yend = 1) +
  annotate("segment", x = 1, xend = 1, y = 0, yend = 1) +
  annotate("segment", x = 0, xend = 1, y = 0, yend = 0) +
  annotate("segment", x = 0, xend = 1, y = 0.2, yend = 0.2) +
  annotate("segment", x = 0, xend = 1, y = 0.4, yend = 0.4) +
  annotate("segment", x = 0, xend = 1, y = 0.6, yend = 0.6) +
  annotate("segment", x = 0, xend = 1, y = 0.8, yend = 0.8) +
  annotate("segment", x = 0, xend = 1, y = 1, yend = 1) +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank()) +
  ggtitle("2D - 25 Regions")

grid.arrange(p1,p2,ncol=2)

```

```{r echo = FALSE}
#CHUNK 3B
scatterplot3d(df$x1, df$x2, df$x3, xlab = "", ylab = "", zlab = "", lab = c(4,4), lab.z = 4, main = "3D - 125 Regions")
```


Consider a simple example of an exponential curve in two dimensional space.

```{r echo=FALSE}
#CHUNK 4A
x1 <- runif(100, 1,10)
x2 <- exp(x1)
df <- data.frame(x1 = x1, x2 = x2)

p1 <- ggplot(data = df, aes(x = x1, y = x2)) + geom_point()
p1
```

CHUNK 4B checks out a log transform.

```{r echo = FALSE}
#CHUNK 4B
df$logx2 <- log(df$x2)
p1 <- ggplot(data = df, aes(x = x1, y = logx2)) + geom_point()
p1
```

CHUNK 5 provides an example of simple classification problem.

```{r echo = FALSE}
#CHUNK 5
df <- data.frame(x1 = c(1,2,1.5,3,3.5,3.5),
                 x2 = c(2,1.5,3,3,3.5,3.9),
                 class = c("A", "A", "A", "B", "B", "B"))

p1 <- ggplot(data = df, aes(x = x1, y = x2, color = class)) +
      geom_point(size = 5) +
      scale_x_continuous(limits = c(0.5,4)) +
      scale_y_continuous(limits = c(0.5,4)) +
      annotate("segment", x = 1.5, xend = 3.5, y = 4, yend = 1, size = 2)
p1

```

Run CHUNK 6 to see the data for a more complex classification problem. You will note that the formula for separating A and B makes it clear that a circle is the best divider. But you wouldn't know that if you only had the data and some noise had been added.

```{r echo = FALSE}
#CHUNK 6
x1 <- runif(1000,-1,1)
x2 <- runif(1000,-1,1)
class <- ifelse(sqrt(x1^2 +x2^2) < 0.5, "A","B")

df <- data.frame(x1 = x1, x2 = x2, class = class)

p1 <- ggplot(data = df, aes(x = x1, y = x2, color = class)) +
      geom_point() 
p1
```

Run CHUNK 7 to see the value of a feature for this problem.

```{r echo = FALSE}
#CHUNK 7
df$dist_from_0 <- sqrt(df$x1^2 + df$x2^2)
      
p2 <- ggplot(data = df, aes(x = 0, y = dist_from_0, color = class)) +
      geom_point(position = "jitter") +
      theme(axis.title.x=element_blank(),
                axis.text.x=element_blank(),
                axis.ticks.x=element_blank()) +
      annotate("segment", x = -0.5, xend = 0.5, y = 0.5, yend = 0.5)

p2

library(scatterplot3d)
p3 <- scatterplot3d(df$x1, df$x2, df$dist_from_0, color = ifelse((df$class == "B"), "blue", "red"))

```

Trivial example of a feature being equal to an input variable.

```{r echo = FALSE} 
#CHUNK 8
df <- data.frame(variable_Income = c(100000, 130000, 95000), feature_Income = c(100000, 130000, 95000))
df
```

An example of converting dates (input variable) to features, such as day of the week or weekday/weekend.

```{r echo = FALSE}
#CHUNK 9
df <- data.frame(variable_date = c("1/8/2017", "12/7/2016", "12/17/2016"),
                 feature_day = c("Sunday", "Wednesday", "Saturday"),
                 feature_dayType = c("Weekend", "Weekday", "Weekend"))
df
```
   
An example of creating more than one feature from a single variable.

```{r echo = FALSE}
#CHUNK 10
df <- data.frame(variable_occupation = c("Pediatric Nurse", "Neurosurgeon", "Carpenter"),
                 feature_occupation_medical = c("TRUE", "TRUE", "FALSE"),
                 feature_occupation_labor = c("FALSE", "FALSE", "TRUE"))
df
```
   
An example of combining multiple variables into a single factor.
    
```{r}
#CHUNK 11
df <- data.frame(variable_sex = c("Female", "Male", "Male"),
                 variable_smoker = c("Non-Smoker", "Non-Smoker", "Smoker"),
                 feature_sex_smoker = c("Female Non-Smoker", "Male Non-Smoker", "Male Smoker"))
df
```

The next chunks are part of a lengthy exercise using the SOA mortality data that you have previously downloaded. This is a large dataset and may take some time to load.

1. Load and summarize the data

This code also creates the target varaible and eliminates the rows in which the target is undefined. You may want to save it with a new name for future use, particularly if you do not complete this session in one sitting. There is code for this in the final chunk in this file.

```{r}
#CHUNK 12
data.mortality <- read.csv("soa_mortality_data.csv")
data.mortality$actual_q <- data.mortality$actual_face / data.mortality$exposure_face
data.mortality <- subset(data.mortality, exposure_face > 0)
nrow(data.mortality)
summary(data.mortality)
```

2. Data transformations

CHUNK 13 has some suggested transformations. Consider adding others as you develop potential features for this dataset.

```{r}
#CHUNK 13
# Log transform example
data.mortality$duration_log <- log(data.mortality$duration)

# Normalization example

data.mortality$issage_norm <- (data.mortality$issage - min(data.mortality$issage)) / (max(data.mortality$issage - min(data.mortality$issage)))

# Binned versions of variables example

data.mortality$issage_bin10 <- cut(data.mortality$issage, 10)
data.mortality$issage_bin20 <- cut(data.mortality$issage, 20, labels = FALSE) # Note the difference between using labels or not

```

3. Binarization
Use CHUNK 14 to create binarized features for all categorical variables except for state.
    
CHUNK 15 provides a sample solution.
    
```{r}
#CHUNK 14

## Your code here

```

CHUNK 15 has a sample solution.

```{r}
#CHUNK 15
library(caret)

# List the variables we want to binarize
vars.bin <- c("sex", "smoker", "prodcat", "region", "distchan", "uwkey", "uwtype", "resind_ind")

# dummyVars is not compatible with factors
for (var in vars.bin) {
  data.mortality[, var] <- as.character(data.mortality[, var])
}

# Binarize variables
binarizer <- caret::dummyVars(paste("~", paste(vars.bin, collapse = "+")) , data = data.mortality, fullRank = F)
data.mortality <- cbind(data.mortality,
                        data.frame(predict(binarizer, newdata = data.mortality)))

```

4. Compound variables

CHUNK 16 has a sample solution for creating two compound variables.
  
```{r}
#CHUNK 16
data.mortality$smoker_sex <- paste(data.mortality$smoker, data.mortality$sex, sep = "_")
data.mortality$smoker_sex_uwtype <- paste(data.mortality$smoker, data.mortality$sex, data.mortality$uwtype, sep = "_")

```

CHUNK 17 has space to create additional variables and also to save the revised file under a new name. Note that future module sections may use the original dataset name and you may wish to modify the code to load this one.

```{r}
#CHUNK 17
#Add code to create any desired adiditonal features



write.csv(data.mortality, file="mortality data with features.csv")
```

